Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-09-02 manu-draft.Rmd
more discussion draft

The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.  Although the seagrass community did not change significantly across the study period, long-term impacts may be observed.  Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that persisted over a year and half in southwest Florida.  These changes may have resulted from a reduction in water clarity from heavy concentrations of *K. brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.  Species shifts from macroalgal  to seagrass dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality of *Caulerpa* dominance relative to seagrasses.  General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communites.  Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Managagement District.  From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.  These broader trends suggest drivers acting at a regional scale are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.  Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
* Potential long-term impacts TBD
* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
* Risk of decline (IRL ex.), regression of past progress

>>>>

The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.  Although the seagrass community did not change significantly across the study period, long-term impacts may be observed.  Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that persisted over a year in southwest Florida.  These changes may have resulted from a reduction in water clarity from heavy concentrations of *K. brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.  Species shifts from macroalgal  to seagrass dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality of *Caulerpa* dominance relative to seagrasses.  General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities.  Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Management District.  From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.  These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.  Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

Potential short and long-term impacts to Tampa Bay resources beyond water quality, macroalgae, and seagrasses may also have occurred.  Fisheries resources have undoubtedly been impacted by *K. brevis* blooms in June and July as documented by the fish kill reports (Figure \@ref(fig:redtide)).  Quantitative data on changes in abundance and diversity of fish species in Tampa Bay following the red tide event are forthcoming. Routine sampling by the Fisheries Independent Monitoring Program of the FWC occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information on species richness and abundance for the major segments of the bay. @Schrandt21 developed a Tampa Bay Nekton Index to assist with fisheries management strategies for the bay and demonstrated a sensitivity of the index to a significant red tide event in lower Tampa Bay in 2005. Given the observed red tide concentration in 2021 and the magnitude of reported fish kills, a similar response in the fish community as reflected by the TBNI is expected.  Of note is that fisheries resources improved quickly following surveys after 2005, although temporary recreational or commercial closures may need to be implemented to allow the fish community to recover.  As of writing, catch and release only limits have been extended through the fall for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*).  @SWitzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following dishcarge to the subembayment and compared the results to fisheries data for years prior.  Fish community structure and species composition did not differ compared to the pre-impact period. From 2003 to 2007, an estimated 1.9 billion gallons was discharged to Bishop Harbor, where various treatment methods were used to mitigate potential impacts of introducing acidic and nutrient-rich process water to surface waters. @Garrett11 noted bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and dense diatom blooms, in Bishop Harbor during this time.

Additional monitoring was conducted for benthic communities to sample sediment quality at Port Manatee and surrounding waters in April 2021.  Biological surveys were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthic community based on the quality of the sediments they inhabit [@tbep0106]. Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions. Results from initial sampling suggest a mix of conditions with dominated by "fair" and "good" samples near Port Manatee reflecting the high spatial variability of macroinvertebrate communities in estuarine sediments.  Comparison of the initial 2021 samples to historical benthic conditions using TBBI scores for comparison suggested relatively consistent sediment quality back to earlist sampline in 1993.  Additional sampling is planned at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality. Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their abundances in the bay and use of bay resources.  Detailed baseline information and data relative to 2021 are lacking for these species relative to the primary data reported herein.  At least one [report](https://myfwc.com/research/manatee/rescue-mortality-response/statistics/mortality/2021/) to FWC of a manatee mortality in Tampa Bay at the peak of the red tide bloom in July was recorded, although cause of death was undetermined. Regardless, it is reasonable to assume toxicity of *K. brevis* can have a negative effect on regional populations of large mammals.  This is of particular concern given the recent report of manatee mass die-offs on the east coast of Florida related to seagrass losses.

* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
* Risk of decline (IRL ex.), regression of past progress
